Temporal Continuity and the Judgment of Actual Causation

### 2021-02-01 docs/resources/CogSci_paper/paper.tex
fit to 6 pages

\newcommand{\ev}[2]{$#1_#2$}

@@

Psychological accounts of actual causation aim to characterize which of multiple causes of an event is singled out as the primary cause. We present one such account called the continuity account. The continuity account treats events as changes of state in continuous time and traces a sequence of stage changes backwards through time from an event to its primary cause. The account is broadly compatible with the physical process view of causation, and we test it by asking people to identify the primary cause of events occurring in simple physical systems. An initial experiment confirms that root causes are more likely to be chosen as primary causes than are immediate causes. A second experiment demonstrates that root causes that have temporal continuity with the effect are preferred even when probability raising accounts would predict otherwise. The results of both experiments are consistent with the continuity account, and suggest that inferences about changes of state in continuous time may underpin an important class of actual causation judgments.

@@

Although we focus on processes unfolding in continuous time, it will be convenient to divide up the temporal dimension into intervals brief enough that there is at most one event per interval. The effect to be explained is an event that happens within one of these intervals: for example, the GCM's change of state from inactive to active. Given this setup, our theory can be formulated using a procedure that starts with the effect and steps backwards through time until the actual cause of the effect is identified.

@@

The physical process account is similarly inconclusive. Both root and immediate causes are connected by a physical process to the effect, and there seems to be no clear reason for preferring one to the another. If we focus only on root causes, again there is no clear preference between \ev{R}{C} and \ev{R}{D}. After the activation of \ev{I}{D}, one can think of this detector as continuously sending activation towards the GCM which only ``unlocks`` the detector once \ev{I}{C} is also active. It follows that both \ev{R}{C} and \ev{R}{D} are connected to the effect by physical processes.

@@

To support these two conclusions we ran two Bayesian mixed effects regression models using the \emph{brms} package in R~\cite{brms} with default priors\footnote{By default brms uses improper flat priors on the coefficients of fixed effects, and a half student-t prior with 3 degrees of freedom, a location of 0 and a scale of 2.5 on the intercept and on the standard deviations of the random effects.}. For the first model, each response was coded as root or immediate, and all responses that could not be classified in this way (including all responses for short sequences) were discarded. We then ran a logistic regression in which stativity (i.e.\ state vs event) and network length (long vs medium) were included as predictors of a binary dependent variable (root vs immediate). We included a random intercept for participant to allow for statistical dependencies between multiple responses from the same participant. The posterior mean of the intercept variable was 10.54 and the 95\% credible interval was $[2.30, 26.54]$, which supports the conclusion that root causes were chosen more often than immediate causes. The credible intervals for the other coefficients both included 0 (state: -2.75, $[-10.03, 1.14]$ ; long: 1.85, $[-1.78, 7.33]$), suggesting that the preference for root causes was not strongly affected by either stativity or network length.

The second analysis was very similar except that the binary dependent variable now indicated whether participants chose a cause on the continuous branch or the delayed branch. The posterior mean of the intercept variable was 2.42 (95\% credible interval $[0.33, 4.83]$), indicating that the continuous branch was chosen more often than the delayed branch. The credible intervals for the state variable also excluded 0 (1.39, $[0.32, 2.51]$), suggesting that state sequences led to a stronger preference for the continuous branch than did event sequences.    Credible intervals for both length variables included 0 (medium: 0.17, $[-0.91,1.31]$ ; long: 0.11, $[-1.01,1.21]$, where the reference level is small), suggesting that the preference for the continuous branch was not affected by network length.

@@

\textbf{Results}.

Network orientation (left or right) and position of the curved branch (top or bottom) had no effect and we collapsed across these variables.  Figure~\ref{fig:expt2}c summarizes the results for the three orders listed in Figures \ref{fig:expt2}a and \ref{fig:expt2}b.  Consistent with Experiment 1,  \ev{R}{C} is preferred over \ev{R}{D} given the order \{\ev{R}{D},\ev{I}{D},\ev{R}{C}\}, and a similar effect was found for the \{\ev{R}{D},\ev{R}{C},\ev{I}{D}\} sequences. The critical finding is that \ev{R}{C} is also preferred for the \{\ev{R}{C},\ev{R}{D},\ev{I}{D}\} sequences, even though the probability raising account makes the opposite prediction. When referring to the orders we have dropped the fourth event because this event is always \ev{I}{C}.

To further analyze the data we used mixed effects models analogous to the two described for Experiment 1. The first analysis used a logistic regression in which order (i.e.\ the temporal order of the key events) was included as a predictor of a binary dependent variable (root or immediate).  The posterior mean of the intercept variable was 5.75 and the 95\% credible interval $[2.44, 10.52]$, which supports the conclusion that root causes were chosen more often than immediate causes. The credible intervals for both order variables included 0 (\{\ev{R}{D},\ev{R}{C},\ev{I}{D}\}: 2.11, $[-1.53,1.21]$; \{\ev{R}{D},\ev{I}{D},\ev{R}{C}\}: -0.38 $[-1.78,0.90]$, where the reference level was \{\ev{R}{C},\ev{R}{D},\ev{I}{D}\}), suggesting that the preference for root causes was fairly consistent across the three orders.

The second and more critical analysis used a binary dependent variable that indicated whether participants chose a cause on the continuous branch or the delayed branch.  The posterior mean of the intercept variable was 2.0 (95\% credible interval $[1.4, 2.65]$), indicating that the continuous branch was chosen more often than the delayed branch.  The credible intervals for both order variables included 0 ( \{\ev{R}{D},\ev{R}{C},\ev{I}{D}\}: -0.38, $[-0.95,0.21]$ ; \{\ev{R}{D},\ev{I}{D},\ev{R}{C}\}: -0.46, $[-1.03,0.12]$) suggesting that the preference for the continuous branch was of similar strength across all three orders. 
Relative to Experiment 1, the frequency with which \ev{I}{C} is chosen has increased in Experiment 2. This difference may reflect the increased difficulty of Experiment 2. When both branches of a dual branch structure are simultaneously active, keeping track of both root causes and the order in which they occurred is relatively challenging, which may lead some participants to fall back on the simple strategy of choosing the immediate cause that directly precedes the effect.

@@

\bibliographystyle{apacite}

\setlength{\bibleftmargin}{.125in}
\setlength{\bibindent}{-\bibleftmargin}

>>>>

\newcommand{\ev}[2]{$#1_#2$}
\renewcommand\bibliographytypesize{\footnotesize}

@@

Psychological theories of actual causation aim to characterize which of multiple causes of an event is singled out as the primary cause. We present one such theory called the continuity account of actual causation. The continuity account treats events as changes of state in continuous time and traces a sequence of stage changes backwards through time from an event to its primary cause. The account is broadly compatible with the physical process view of causation, and we test it by asking people to identify the primary cause of events occurring in simple physical systems. An initial experiment confirms that root causes are more likely to be chosen as primary causes than are immediate causes. A second experiment demonstrates that root causes that have temporal continuity with the effect are preferred even when probability raising accounts would predict otherwise. The results of both experiments are consistent with the continuity account, and suggest that inferences about changes of state in continuous time may underpin an important class of actual causation judgments.

@@

Although we focus on processes unfolding in continuous time, it will be convenient to divide up the temporal dimension into intervals brief enough that at most one event occurs per interval. The effect to be explained is an event that happens within one of these intervals: for example, the GCM's change of state from inactive to active. Given this setup, our theory can be formulated using a procedure that starts with the effect and steps backwards through time until the actual cause of the effect is identified.

@@

The physical process account is similarly inconclusive. Both root and immediate causes are connected by a physical process to the effect, and there seems to be no clear reason for preferring one to the another. If we focus only on root causes, again there is no clear preference between \ev{R}{C} and \ev{R}{D}. After the activation of \ev{I}{D}, one can think of this detector as continuously sending activation towards the GCM which only ``unlocks'' the detector once \ev{I}{C} is also active. It follows that both \ev{R}{C} and \ev{R}{D} are connected to the effect by physical processes.

@@

To support these two conclusions we ran two Bayesian mixed effects regression models using the \emph{brms} package in R~\cite{brms} with default priors\footnote{By default brms uses improper flat priors on the coefficients of fixed effects, and a half student-t prior with 3 degrees of freedom, a location of 0 and a scale of 2.5 on the intercept and on the standard deviations of the random effects.}. For the first model, each response was coded as root or immediate, and all responses that could not be classified in this way (including all responses for short sequences) were discarded. We then ran a logistic regression in which stativity (i.e.\ state vs event) and network length (long vs medium) were included as predictors of a binary dependent variable (root vs immediate). We included a random intercept for participant to allow for statistical dependencies between multiple responses from the same participant. The posterior mean of the intercept variable was 10.8 and the 95\% credible interval was $[2.4, 26.8]$, which supports the conclusion that root causes were chosen more often than immediate causes. The credible intervals for the other coefficients both included 0 (state: -2.7, $[-9.2, 1.2]$ ; long: 2.1, $[-1.7, 9.1]$), suggesting that the preference for root causes was not strongly affected by either stativity or network length.

The second analysis was very similar except that the binary dependent variable now indicated whether participants chose a cause on the continuous branch or the delayed branch. The posterior mean of the intercept variable was 1.7 but the 95\% credible interval included zero ($[-0.07, 3.8]$), suggesting only weak support for the conclusion that the continuous branch was chosen more often than the delayed branch. The credible intervals for the state variable excluded 0 (1.0, $[0.05, 1.9]$), suggesting that state sequences led to a stronger preference for the continuous branch than did event sequences.    Credible intervals for both length variables included 0 (medium: 0.4, $[-0.5,1.4]$ ; long: 0.5, $[-0.5,1.4]$, where the reference level is small), suggesting that the preference for the continuous branch was not strongly affected by network length.

@@

\textbf{Results}.  
Network orientation (left or right) and position of the curved branch (top or bottom) had no effect and we collapsed across these variables.  Figure~\ref{fig:expt2}c summarizes the results for the three orders listed in Figures \ref{fig:expt2}a and \ref{fig:expt2}b.  Consistent with Experiment 1,  \ev{R}{C} is preferred over \ev{R}{D} given the order \{\ev{R}{D},\ev{I}{D},\ev{R}{C}\}, and a similar effect was found for the \{\ev{R}{D},\ev{R}{C},\ev{I}{D}\} sequences. The critical finding is that \ev{R}{C} is also preferred for the \{\ev{R}{C},\ev{R}{D},\ev{I}{D}\} sequences, even though the probability raising account makes the opposite prediction. When referring to the orders we have dropped the fourth event because this event is always \ev{I}{C}.

To further analyze the data we used mixed effects models analogous to the two described for Experiment 1. The first analysis used a logistic regression in which order (i.e.\ the temporal order of the key events) was included as a predictor of a binary dependent variable (root or immediate).  The posterior mean of the intercept variable was 5.2 and the 95\% credible interval $[2.0, 9.4]$, which supports the conclusion that root causes were chosen more often than immediate causes. The credible intervals for both order variables included 0 (\{\ev{R}{D},\ev{R}{C},\ev{I}{D}\}: -0.2, $[-1.4,1.2]$; \{\ev{R}{D},\ev{I}{D},\ev{R}{C}\}: -0.4 $[-1.7,0.9]$, where the reference level was \{\ev{R}{C},\ev{R}{D},\ev{I}{D}\}), suggesting that the preference for root causes was fairly consistent across the three orders.

The second and more critical analysis used a binary dependent variable that indicated whether participants chose a cause on the continuous branch or the delayed branch.  The posterior mean of the intercept variable was 1.9 (95\% credible interval $[1.3, 2.5]$), indicating that the continuous branch was chosen more often than the delayed branch.  The credible intervals for both order variables included 0 ( \{\ev{R}{D},\ev{R}{C},\ev{I}{D}\}: -0.4, $[-1,0.2]$ ; \{\ev{R}{D},\ev{I}{D},\ev{R}{C}\}: -0.5, $[-1,0.1]$) suggesting that the preference for the continuous branch was of similar strength across all three orders.

Relative to Experiment 1, the frequency with which \ev{I}{C} is chosen has increased in Experiment 2. This difference may reflect the increased difficulty of Experiment 2. When both branches of a dual branch structure are simultaneously active, keeping track of both root causes and the order in which they occurred is relatively challenging, which may lead some participants to fall back on the simple strategy of choosing the immediate cause that directly precedes the effect.

@@

\bibliographystyle{apacite}
\setlength{\bibleftmargin}{.125in}
\setlength{\bibindent}{-\bibleftmargin}
